Cooperative Kernels: Safe Multitasking for Blocking Algorithms on GPUs

### 2016-11-09 write-ups/PLDI/paper.tex
Revisions discussed with Sreepathi.

@@

\paragraph{The needs of irregular data-parallel algorithms}
Acceleration of general-purpose computations on graphics processing
units (GPUs) has tended to focus on \emph{regular} data-parallel
algorithms, for which work to be processed can be evenly split between
the workgroups that execute a GPU kernel ahead of time.  However, many
interesting data-parallel algorithms are \emph{irregular}: the amount
of work to be processed is unknown ahead of time and may be determined
by the computation itself.  There is growing interest in accelerating irregular
data-parallel algorithms on GPUs, particularly algorithms that
traverse and manipulate graphs, with several recent notable
successes~\cite{...}.

Irregular algorithms usually require \emph{blocking synchronization}
between workgroups, to balance load and communicate intermediate
results.  For example, work stealing algorithms require each workgroup
to maintain a queue, which is typically protected by a mutex to enable
stealing by other workgroups.  As another example, breadth-first graph
traversal uses a frontier-based approach that requires a barrier
synchronization between all workgroups to achieve consensus that
exploration of a frontier has been accomplished.  \ADComment{Perhaps
  make the point that inter-workgroup synchronization is possible in
  very rare cases without blocking, e.g. threadfence reduction.}

To operate correctly, a blocking concurrent algorithm requires
\emph{fair} scheduling of threads.  In the context of irregular
data-parallel algorithms accelerated on a GPU, this translates to
requiring fair scheduling of workgroups.  Without fairness, a
workgroup may fail to make progress, leading to starvation.  For
example, if workgroup $A$ holds a mutex and workgroup $B$ is
spin-waiting for the mutex to be released, an unfair scheduler can
forever schedule $B$, so that $A$ never releases the mutex.
Similarly, an unfair scheduler can induce starvation if workgroup $A$
is at an inter-workgroup barrier awaiting the arrival of other
workgroups, by allowing A to spin forever so that the other workgroups
do not progress to the barrier.

\paragraph{Occupancy-bound execution}
Current GPU programming models---CUDA~\cite{...}, OpenCL~\cite{...}
and HSA~\cite{...}---specify very few constraints on the manner in
which workgroups are scheduled.  In particular, they do \emph{not}
specify that workgroups must be fairly scheduled, and implementations
of these programming models do not provide fair scheduling in
practice.  Roughly speaking, each workgroup executing a GPU kernel is
mapped to a hardware \emph{compute unit}.  If a kernel is executed
with more workgroups than there are compute units then at least two
workgroups must share a single compute unit.  The simplest way to
handle this is via an \emph{occupancy-bound} execution model, whereby
once a workgroup has commenced execution on a compute unit (is has
become \emph{occupant}), the workgroup has exclusive access to the
compute unit until it has finished execution, after which a subsequent
workgroup can be bound to the compute unit.  Experiments suggest that
the occupancy-bound execution model is widely implemented by today's
GPU devices and drivers~\cite{...}, and it is evidently a simple model
for device vendors to implement efficiently.

The occupancy-bound execution model does not guarantee fair scheduling
between workgroups: if all compute units are occupied then a
not-yet-occupant workgroup will not be scheduled until some occupant
workgroup completes execution.  Yet the execution model \emph{does}
provide fair scheduling between \emph{occupant} workgroups, because
they are bound to separate compute units that operate in parallel.
Current implementations of irregular data parallel algorithms on GPUs
assume that the target GPU device provides the occupancy-bound
execution model, which they exploit by scheduling no more workgroups
than there are available compute units~\cite{...}.  This works today
because \emph{today}'s GPUs seem to provide the occupancy-bound
execution model.

\paragraph{Resistance to occupancy-bound execution}

Despite its practical prevalence, none of the current GPU programming
models actually mandate occupancy-bound execution.  Further, there are
clear reasons why this model is undesirable from the point-of-view of
GPU vendors and end users.  First, the execution model does not enable
multi-tasking, because a workgroup effectively \emph{owns} a compute
unit until the workgroup has completed execution.  If a long-running
kernel occupies all workgroups, the GPU cannot be used for other
tasks, such as graphics rendering, in the meantime.  Precisely because
of this problem, operating systems employ a GPU \emph{watchdog} to
heuristically detect and kill long-running computations on the GPU
being used for display rendering.  If this watchdog is disabled,
execution of a long-running kernel causes the display to freeze for
the duration of kernel execution.  Second, \emph{energy throttling} is
an important concern for battery-powered devices, such as smartphones
and tablets.  It is desirable for a mobile GPU driver to be able to
power down one or more compute units if the battery level is low or if
the user puts the device into a low power state.  Implementing energy
throttling requires being able to suspend the execution of a
workgroup, power down the compute unit to which the workgroup was
bound, and schedule the workgroup for later execution on a different
compute unit.

Our assessment, informed by discussions with a number of industrial
practitioners who have been involved in the OpenCL and/or HSA
standardisation efforts~\cite{CitePersonalCommunication}, is that
\textbf{(1)} GPU vendors will not commit to the occupancy-bound
execution model that they currently implement, due to the need for
multi-tasking and energy throttling, yet \textbf{(2)} GPU vendors will
not guarantee fair scheduling, due to the runtime overhead associated
with supporting full preemption of workgroups, e.g.\ based on
time-slicing.  Vendors instead wish to retain the essence of the
occupancy-bound execution model, which is simple to implement
efficiently, and provide preemption for the special cases associated
with multi-tasking for high-priority jobs (most crucially graphics)
and energy throttling.  In particular, they do not want to commit to
fair scheduling of a workgroup once it has been preempted.

@@

At a point where the cooperative kernel could benefit from having
additional workgroups join the computation, a workgroup can execute
$\offerfork$.  This indicates to the scheduler that the cooperative
kernel is prepared to proceed with the existing set of workgroups, but
is able to take advantage of having one or more additional workgroups
commencing execution at the $\offerfork$ program point, inheriting the
state of the workgroup that executed $\offerfork$.

The use of $\offerfork$ and $\offerkill$ creates a contract between
the scheduler and (the programmer of the) cooperative kernel, with
both functional and non-functional requirements.  Functionally: the
scheduler must guarantee that the workgroups executing a cooperative
kernel are fairly scheduled; the cooperative kernel must be robust to
workgroups leaving and joining the computation when the $\offerkill$
and $\offerfork$ functions, respectively, are called.
Non-functionally: a cooperative kernel must ensure that $\offerkill$
is executed frequently enough that the scheduler can accommodate
soft-real time constraints, e.g.\ allowing a smooth frame-rate for
graphics or allowing compute units to be powered down quickly when
required; the scheduler should respond to $\offerkill$ and
$\offerfork$ calls in a manner that avoids under-utilisation of
hardware resources by the cooperative kernel, e.g.\ the scheduler
should not kill workgroups unless they are required for another
purpose.

Cooperative kernels have several appealing properties.  First, they
accommodate the needs of blocking algorithms, including irregular
data-parallel algorithms: such algorithms can be implemented naturally
as cooperative kernels \ADComment{boast that we did a good set}, so
that fair scheduling between workgroups is guaranteed.  Second, they
have no impact on the development of regular (non-cooperative
kernels): these can be programmed exactly as they are now.  Third, the
programming model is backwards compatible: a non-blocking algorithm
implementation that works on today's GPU platforms can be upgraded to
use our language extensions for cooperative scheduling, and will
behave exactly as it does at present if the new $\offerkill$ and
$\offerfork$ functions are simply ignored.  Fourth, the programming
model is easy to implement on top of the occupancy-bound execution
model that current GPU platforms provide; we present a
proof-of-concept implementation on top of OpenCL 2.0 that requires no
special-purpose hardware or driver support.  Fifth, our results
suggest that cooperative kernels can provide efficient multi-tasking
of cooperative and non-cooperative tasks: our experiments across a
range of GPUs from two vendors show very good response times for
scheduling short, non-cooperative kernels concurrently with
long-running cooperative kernels.  These performance results are with
respect to our proof-of-concept, portable implementation; a vendor
specific implementation of cooperative kernels could likely yield
better performance still through device-specific driver and compiler
support.

@@

@@

A regular OpenCL kernel can be executed by a multi-dimensional grid of
workgroups, called an ND-range.  In our current proposal, the ND-range
for a cooperative kernel must be one-dimensional.  \ADComment{Say why?
  And mention that none of the applications that we deem would benefit
  from cooperative kernels need a multi-dimensional ND-range.}

@@

A variable declared in the root scope of the cooperative kernel can
optionally be annotated with a new \texttt{restore} qualifier.  By
annotating a variable with \texttt{restore}, the programmer indicates
that when new workgroups are spawned via $\offerkill$, the threads in
these workgroups should obtain a well-defined initial value for the
variable, obtained from the workgroup that executed $\offerkill$.  We
detail the semantics for this when we describe $\offerkill$ below.

Only private memory variables can be annotated with \texttt{restore},
which may include pointers to local or global memory so long as the
pointer variables themselves are stored in private memory.  In
particular, variables and arrays in local memory cannot be annotated
with \texttt{restore}.  \ADComment{The preceding is what we have
  implemented, but actually it is pretty clear how we could handle
  local memory restoration if we wanted to; it's just that none of our
  examples needed it.}

@@

@@

\item for each variable $v$ annoated with \texttt{restore}, every new
  thread's copy of $v$ is initialised to the value that thread 0 in
  workgroup $m$ held for $v$ at the point of the $\offerfork$ call.

@@

The second exception, for restored variables, can be stated more
formally as follows.  Let $\mathit{WG}[x][y].v$ denote the value
stored in variable $v$ of thread $y$ in workgroup $x$.  Then for each
new workgroup $i \in [M, M+k-1]$ and each thread $j$ in workgroup $i$,
$\mathit{WG}[i][j].v$ is initialised to $\mathit{WG}[m][0].v$ for each
variable $v$ annotated with \texttt{restore}.

Notice that $k=0$ is always a legitimate choice for the number of
workgroups to be spawned by an $\offerfork$ call, and if the number of
active workgrouups $M$ is equal to the workgroup limit $N$, $k=0$ is
guaranteed.

@@

Many irregular algorithms require synchronisation across workgroups.
This can be achieved via a \emph{global barrier}, but a global barrier
is not provided as a primitive in OpenCL, CUDA or HSA.  This is
because, as discussed in Section~\ref{ }, an inter-workgroup barrier
requires fair scheduling of workgroups.  As a result, past works have
either split computation into multiple separate kernels, so that the
end of one kernel and the start of the next kernel implicitly provides
a global barrier~\cite{ }, or have assumed the occupancy-bound
execution model to build a global barrier that works reliably on
today's GPU platforms.

@@

The $\globalbarrier()$ primitive can be implemented directly using the
atomic operations provided by the OpenCL 2.0 memory model~\cite{...},
but the implementation is involved, hence why we include this function
as a primitive.  As noted above, the $\resizingglobalbarrier()$
primitive is, semantically, synactic sugar for a sequence of other
primitive calls.  We include it as a primitive because (a) we have
found that global barriers often provide a natural point for
cooperation with the scheduler, and (b) identifying a known point at
which workgroups are required to synchronize and at which workgroups
can be safely killed and forked enables a custom implementation of
$\resizingglobalbarrier()$ that provides the semantics described
above, but is significantly more efficient than the above sequence of
calls.  We discuss our efficient implementation of
$\resizingglobalbarrier$ in detail in Section~\ref{...}.

\subsection{Backwards compatibility}

\ADComment{Make the point here that if an irregular algorithm is
  written as a cooperative kernel then it will execute just as safely
  as it does today on current hardware if: (1) the \texttt{restore}
  annotation and $\offerkill$/$\offerfork$ calls are defined away to
  nothing, (2) $\resizingglobalbarrier$ is redefined to
  $\globalbarrier$, (3) a standard implementation of $\globalbarrier$
  that assumes the occupancy-bound execution model is provided.  If the kernel uses blocking features---the global barrier, mutexes, etc., then it may not execute safely due to the possibility of unfair scheduling, but it will execute no less safely than it would if the cooperative kernels style were not used.}

@@

\lstset{basicstyle=\tt}
\begin{lstlisting}
  if($\getgroupid{0}$ != 0) $\offerkill$;
\end{lstlisting}
\lstset{basicstyle=\scriptsize\tt}

@@

The main impact of this relates to partitioning of data to be
processed.

\ADComment{Tyler, can we give an example of a data processing loop
  that would have to be changed?}

@@

In this case, because the resizing barrier is a kernel-level function,
the statements executed by the cooperative kernel can be divided into
\emph{resizing barrier intervals}: a resizing barrier interval starts
at a resizing barrier or the kernel entry point, and ends at a
resizing barrier or a kernel exit point.  At any moment of execution,
all threads are guaranteed to be executing within the same resizing
barrier interval.  As a result, the number of workgroups executing the
kernel is guaratneed to remain unchanged within each resizing barrier
interval, so the result returned by $\getnumgroups{0}$ is guaratneed
to be identical on repeated calls during such an interval.  This
guarantee may be exploited for efficiently when writing a cooperative
kernel that uses global barriers but does not call $\offerfork$ and
$\offerkill$ directly.

@@

@@

@@

>>>>

\newcommand{\restore}{\texttt{restore}}

@@

\ADComment{Think about whether ``restore'' is best name.}

\ADComment{Address issue of placing constructs manually --
  transactional semantics and barriers.  Argue why it's easy to do this.}

\ADComment{Mention dynamic parallelism ane explain why it's no good.}

\paragraph{The needs of irregular data-parallel algorithms}
Acceleration of general-purpose computations on graphics processing
units (GPUs) has tended to focus on \emph{regular} data-parallel
algorithms, for which work to be processed can be evenly split between
the workgroups that execute a GPU kernel ahead of time.  However, many
interesting data-parallel algorithms are \emph{irregular}: the amount
of work to be processed is unknown ahead of time and may be determined
by the computation itself.  There is growing interest in accelerating irregular
data-parallel algorithms on GPUs, particularly algorithms that
traverse and manipulate linked data structures, with several recent notable
successes~\cite{...}.

Irregular algorithms usually require \emph{blocking synchronization}
between workgroups, to balance load and communicate intermediate
results.  For example, many graph algorithms employ a level-by-level strategy, requiring a global barrier between each level.
As another example, work stealing algorithms require each workgroup
to maintain a queue, which is typically protected by a mutex to enable
stealing by other workgroups.  \ADComment{Perhaps
  make the point that inter-workgroup synchronization is possible in
  very rare cases without blocking, e.g. threadfence reduction.}

To operate correctly, a blocking concurrent algorithm requires
\emph{fair} scheduling of threads.  In the context of irregular
data-parallel algorithms accelerated on a GPU, this translates to
requiring fair scheduling of workgroups.  \ADComment{Shorten} Without fairness, a
workgroup may fail to make progress, leading to starvation.  For
example, if workgroup $A$ holds a mutex and workgroup $B$ is
spin-waiting for the mutex to be released, an unfair scheduler can
forever schedule $B$, so that $A$ never releases the mutex.
Similarly, an unfair scheduler can induce starvation if workgroup $A$
is at an inter-workgroup barrier awaiting the arrival of other
workgroups, by allowing A to spin forever so that the other workgroups
do not progress to the barrier.

\paragraph{Occupancy-bound execution}
\ADComment{Add footnote to acknowledge multiple wgs per cu}
\ADComment{Say that current GPU schedulers are not required to be fair.}
Current GPU programming models---CUDA~\cite{...}, OpenCL~\cite{...}
and HSA~\cite{...}---specify very few constraints on the manner in
which workgroups are scheduled.  In particular, they do \emph{not}
specify that workgroups must be fairly scheduled, and implementations
of these programming models do not provide fair scheduling in
practice.  Roughly speaking, each workgroup executing a GPU kernel is
mapped to a hardware \emph{compute unit}.  If a kernel is executed
with more workgroups than there are compute units then at least two
workgroups must share a single compute unit.  The simplest way to
handle this is via an \emph{occupancy-bound} execution model, whereby
once a workgroup has commenced execution on a compute unit (it has
become \emph{occupant}), the workgroup has exclusive access to the
compute unit until it has finished execution, after which a subsequent
workgroup can be bound to the compute unit.  Experiments suggest that
the occupancy-bound execution model is widely implemented by today's
GPU devices and drivers~\cite{...}, and it is evidently a simple model
for device vendors to implement efficiently.

The occupancy-bound execution model does not guarantee fair scheduling
between workgroups: if all compute units are occupied then a
not-yet-occupant workgroup will not be scheduled until some occupant
workgroup completes execution.  Yet the execution model \emph{does}
provide fair scheduling between \emph{occupant} workgroups, because
they are bound to separate compute units that operate in parallel.
Current implementations of irregular data parallel algorithms on GPUs
assume that the target GPU device provides the occupancy-bound
execution model, which they exploit by scheduling no more workgroups
than there are available compute units~\cite{...}.  This works
because \emph{today}'s GPUs seem to provide the occupancy-bound
execution model.

\paragraph{Resistance to guaranteeing occupancy-bound execution}

Despite its practical prevalence, none of the current GPU programming
models actually mandate occupancy-bound execution.  Further, there are
clear reasons why this model is undesirable from the point-of-view of
GPU vendors and end users.  First, the execution model does not enable
multi-tasking, because a workgroup effectively \emph{owns} a compute
unit until the workgroup has completed execution.  If a long-running
kernel occupies all compute units, the GPU cannot be used for other
tasks, such as graphics rendering, in the meantime.  \ADComment{Could cut or move to later:} Precisely because
of this problem, operating systems employ a GPU \emph{watchdog} to
heuristically detect and kill long-running computations on the GPU
being used for display rendering.  If this watchdog is disabled,
execution of a long-running kernel causes the display to freeze for
the duration of kernel execution.  Second, \emph{energy throttling} is
an important concern for battery-powered devices, such as smartphones
and tablets.  In the future, it will be desirable for a mobile GPU driver to be able to
power down one or more compute units if the battery level is low or if
the user puts the device into a low power state.  Implementing energy
throttling would require being able to suspend the execution of a
workgroup, power down the compute unit to which the workgroup was
bound, and schedule the workgroup for later execution on a different
compute unit.

Our assessment, informed by discussions with a number of industrial
practitioners who have been involved in the OpenCL and/or HSA
standardisation efforts~\cite{CitePersonalCommunication}, is that
\textbf{(1)} GPU vendors will not commit to the occupancy-bound
execution model that they currently implement, due to the need for
multi-tasking and energy throttling, yet \textbf{(2)} GPU vendors will
not guarantee fair scheduling, due to the runtime overhead associated
with supporting full preemption of workgroups~\cite{ISCAPAPER}.  Vendors instead wish to retain the essence of the
occupancy-bound execution model, which is simple to implement
efficiently, and provide preemption for \ADComment{shorten} the special cases associated
with multi-tasking for high-priority jobs (most crucially graphics)
and energy throttling.  In particular, they do not want to commit to
fair scheduling of a workgroup once it has been preempted.

\ADComment{Kepler preemption support?}

@@

At a point where the cooperative kernel could benefit from having
additional workgroups join the computation, a workgroup can execute
$\offerfork$.  This indicates to the scheduler that the cooperative
kernel is prepared to proceed with the existing set of workgroups, but
is able to take advantage of having one or more additional workgroups
commencing execution at the $\offerfork$ program point.

The use of $\offerfork$ and $\offerkill$ creates a contract between
the scheduler and (the programmer of the) cooperative kernel, with
both functional and non-functional requirements.  Functionally, the
scheduler must guarantee that the workgroups executing a cooperative
kernel are fairly scheduled, while the cooperative kernel must be robust to
workgroups leaving and joining the computation in response to $\offerkill$
and $\offerfork$.
Non-functionally, a cooperative kernel must ensure that $\offerkill$
is executed frequently enough that the scheduler can accommodate
soft-real time constraints, e.g.\ allowing a smooth frame-rate for
graphics or allowing compute units to be powered down quickly when
required.  The scheduler should respond to $\offerkill$ and
$\offerfork$ calls in a manner that avoids under-utilisation of
hardware resources by the cooperative kernel, e.g.\ the scheduler
should not kill workgroups unless they are required for another
purpose.

Cooperative kernels have several appealing properties:

\item They
accommodate the needs of the irregular
data-parallel algorithms that have been successfully accelerated using GPUs: such algorithms can be implemented naturally
as cooperative kernels \ADComment{boast that we did a good set}, so
that fair scheduling between workgroups is guaranteed.

\item They
have no impact on the development of regular (non-cooperative
kernels): these can be programmed exactly as they are now.

\item The
programming model is backwards compatible: a non-blocking algorithm
implementation that works on today's GPU platforms can be upgraded to
use our language extensions for cooperative scheduling, and will
behave exactly as it does at present if the new $\offerkill$ and
$\offerfork$ functions are simply ignored.

\item The programming
model is easy to implement on top of the occupancy-bound execution
model that current GPU platforms provide; we present a
proof-of-concept implementation on top of OpenCL 2.0 that requires no
special-purpose hardware or driver support.

\item Our results
suggest that cooperative kernels can provide efficient multi-tasking
of cooperative and non-cooperative tasks: our experiments across a
range of GPUs from two vendors show very good response times for
scheduling short, non-cooperative kernels concurrently with
long-running cooperative kernels.  These performance results are with
respect to our proof-of-concept, portable implementation; a vendor
specific implementation of cooperative kernels could likely yield
better performance still through device-specific driver and compiler
support.

@@

\ADComment{A natural exit does not decrement $M$.}

@@

\ADComment{Perhaps remove this because a driver-based approach could handle this.}
A regular OpenCL kernel can be executed by a multi-dimensional grid of
workgroups, called an ND-range.  In our current proposal, the ND-range
for a cooperative kernel must be one-dimensional.  \ADComment{Say why?
  And mention that none of the applications that we deem would benefit
  from cooperative kernels need a multi-dimensional ND-range.}

@@

A variable declared in the root scope of the cooperative kernel can
optionally be annotated with a new \restore qualifier.  By
annotating a variable with \restore, the programmer indicates
that when new workgroups are spawned via $\offerfork$, the threads in
these workgroups should obtain a well-defined initial value for the
variable, obtained from the workgroup that executed $\offerfork$.  We
detail the semantics for this when we describe $\offerfork$ below.

\ADComment{Remove?}
Only private memory variables can be annotated with \restore,
which may include pointers to local or global memory so long as the
pointer variables themselves are stored in private memory.  In
particular, variables and arrays in local memory cannot be annotated
with \restore.  \ADComment{The preceding is what we have
  implemented, but actually it is pretty clear how we could handle
  local memory restoration if we wanted to; it's just that none of our
  examples needed it.}

@@

\ADComment{Don't allow zero to be killed.}

\ADComment{Explain that the ``last one dies'' semantics is convenient because it maintains contiguous ids, but other options are possible and may be more efficient sometimes.}

@@

\item for each variable $v$ annoated with \restore, every new
  thread's copy of $v$ is initialised to the value that thread 0 in
  workgroup $m$ held for $v$ at the point of the $\offerfork$ call.

@@

Notice that $k=0$ is always a legitimate choice for the number of
workgroups to be spawned by an $\offerfork$ call, and if the number of
active workgroups $M$ is equal to the workgroup limit $N$, $k=0$ is
guaranteed.

@@

Many irregular algorithms require synchronisation across workgroups.
This can be achieved via a \emph{global barrier}, but a global barrier
is not provided as a primitive in OpenCL, CUDA or HSA.  This is
because, as discussed in Section~\ref{ }, an inter-workgroup barrier
requires fair scheduling of workgroups.

@@

The $\globalbarrier()$ primitive can be implemented directly using the
atomic operations provided by the OpenCL 2.0 memory model~\cite{...},
but the implementation is involved, hence why we include this function
as a primitive.  As noted above, the $\resizingglobalbarrier()$
primitive is, semantically, synactic sugar for a sequence of other
primitive calls.  We include it as a primitive because (a) we have
found that global barriers often provide a natural point for
cooperation with the scheduler, and (b) a custom implementation of
$\resizingglobalbarrier()$ that provides the semantics described
above can be made significantly more efficient than the above sequence of
calls.  We discuss our efficient implementation of
$\resizingglobalbarrier$ in detail in Section~\ref{...}.

@@

\lstset{basicstyle=\tt}
\begin{lstlisting}
  if($\getgroupid{0}$ != 0) $\offerkill$();
\end{lstlisting}
\lstset{basicstyle=\scriptsize\tt}

@@

@@

\ADComment{Shorten sentences.}
In this case, because the resizing barrier is a kernel-level function,
the statements executed by the cooperative kernel can be divided into
\emph{resizing barrier intervals}: a resizing barrier interval starts
at a resizing barrier or the kernel entry point, and ends at a
resizing barrier or a kernel exit point.  At any moment of execution,
all threads are guaranteed to be executing within the same resizing
barrier interval.  As a result, the number of workgroups executing the
kernel is guaratneed to remain unchanged within each resizing barrier
interval, so the result returned by $\getnumgroups{0}$ is guaratneed
to be identical on repeated calls during such an interval.

This guarantee may be exploited by algorithms that peform strided data
processing.  \ADComment{Tyler, can we give an example of a data
  processing loop that would have to be changed?}

@@

\ADComment{Sree mentioned TimeGraph paper.}

@@

\ADComment{Forward reference to experiments with display freezing.}

@@

\ADComment{Generosity: initial number of workgroups, and responding to offer fork.}

\subsection{Alternative Semantic Choices}

\ADComment{Explain here where we could have done things differently.}

\ADComment{What are difficultiese with other alternatives and why we didn't use them.}

\ADComment{Allowing zero to be killed.}

\subsection{Backwards Compatibility}

\ADComment{Make the point here that if an irregular algorithm is
  written as a cooperative kernel then it will execute just as safely
  as it does today on current hardware if: (1) the \restore
  annotation and $\offerkill$/$\offerfork$ calls are defined away to
  nothing, (2) $\resizingglobalbarrier$ is redefined to
  $\globalbarrier$, (3) a standard implementation of $\globalbarrier$
  that assumes the occupancy-bound execution model is provided.  If the kernel uses blocking features---the global barrier, mutexes, etc., then it may not execute safely due to the possibility of unfair scheduling, but it will execute no less safely than it would if the cooperative kernels style were not used.}
